---
title: "Tooth Growth Data Inference"
author: "Baoshi Sun"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

## Overview
In this report, we are going to perform some exploratory analyses on the ToothGrowth data in the R datasets package and try to reach conclusions based on the results of the confidence intervals and hypothesis test.

## Exploratory Analysis
```{r BasisExploratory, echo=FALSE}
data(ToothGrowth)
str(ToothGrowth)
table(ToothGrowth$supp); table(ToothGrowth$dose)
```

We have 60 samples of 3 variables in the dataset, which can be splitted into 2 groups evenly on variable supp (VC and OJ) with 3 amount of doses (0.5, 1, 2) in each group.

```{r plotExploratory, echo=FALSE}
with(ToothGrowth, {
    boxplot(len ~ dose, boxwex = 0.25, at = 1:3 - 0.2,
        subset = supp == "VC", col = "red",
        main = "Tooth Growth under different dose of supp",
        xlab = "dose of supp", ylab = "tooth length",
        xlim = c(0.5, 3.5), ylim = c(0, 35), yaxs = "i")
    boxplot(len ~ dose, add = TRUE, boxwex = 0.25, at = 1:3 + 0.2,
        subset = supp == "OJ", col = "orange")
    legend(2.91, 12.1, c("VC", "OJ"), fill = c("red", "orange"))
})
```

Intuitionally, OJ may lead to more tooth growth than VC given the same dose. And greater dose seems to have greater effect.

## Confidence Intervals on variable supp

```{r suppAnalysis, echo=FALSE}
g_vc <- ToothGrowth$len[ToothGrowth$supp == "VC"]
g_oj <- ToothGrowth$len[ToothGrowth$supp == "OJ"]
n_vc <- length(g_vc); n_oj <- length(g_oj)
m_vc <- round(mean(g_vc), 2); m_oj <- round(mean(g_oj), 2)
s_vc <- round(sd(g_vc) / sqrt(n_vc), 2); s_oj <- round(sd(g_oj) / sqrt(n_oj), 2)

sum_supp <- data.frame(rbind(c("VC", n_vc, m_vc, s_vc), c("OJ", n_oj, m_oj, s_oj)))
names(sum_supp) <- c("supp", "count", "mean", "sd")
sum_supp
```

To verify the idea of OJ having better outcome over VC, we divide the data into two small-sample groups on variable supp. Because the groups are independent, we are using indenpendent group t confidence intervals with equal variance.   

```{r suppTest, echo=FALSE}
t.test(g_oj, g_vc, paired=FALSE, var.equal=TRUE)$conf
t.test(g_oj, g_vc, paired=FALSE, var.equal=TRUE, conf.level = 0.9)$conf
```

Let us construct both 95% and 90% T intervals for supp comparing OJ to VC (substracting in order of OJ - VC). As can be seen, under in confidence level 95% the lower end of the T interval is below zero, whereas the interval is entirely above zero if we set the confidence level to 90%.  

## Hypothesis Testing on variable dose
```{r doseAnalysis, echo=FALSE}
g_dose1 <- ToothGrowth$len[ToothGrowth$dose == 0.5]
g_dose2 <- ToothGrowth$len[ToothGrowth$dose == 1]
g_dose3 <- ToothGrowth$len[ToothGrowth$dose == 2]
n_dose1 <- length(g_dose1); n_dose2 <- length(g_dose2); n_dose3 <- length(g_dose3)
m_dose1 <- round(mean(g_dose1), 2)
m_dose2 <- round(mean(g_dose2), 2)
m_dose3 <- round(mean(g_dose3), 2)
s_dose1 <- round(sd(g_dose1) / sqrt(n_dose1), 2)
s_dose2 <- round(sd(g_dose2) / sqrt(n_dose2), 2)
s_dose3 <- round(sd(g_dose3) / sqrt(n_dose3), 2)

sum_dose <- data.frame(rbind(c(0.5, n_dose1, m_dose1, s_dose1), c(1, n_dose2, m_dose2, s_dose2), c(2, n_dose3, m_dose3, s_dose3)))
names(sum_dose) <- c("dose", "count", "mean", "sd")
sum_dose
```

```{r doseTest, echo=FALSE}
ht_d1 <- t.test(g_dose2, g_dose1, paired=FALSE, var.equal=TRUE, alternative="greater")
al_d1 <- ht_d1$alternative; pv_d1 <- ht_d1$p.value
ht_d2 <- t.test(g_dose3, g_dose2, paired=FALSE, var.equal=TRUE, alternative="greater")
al_d2 <- ht_d2$alternative; pv_d2 <- ht_d2$p.value
```

### Hypothesis Testing 1: dose 1 on does 0.5
* $H_0 : \mu = `r m_dose1`$   
* $H_a : \mu > `r m_dose1`$   
When tested under 95% confidence level, we got p-value `r pv_d1`, which is less than $\alpha = 0.05$. Therefore, we can reject the null hypotheis.  

### Hypothesis Testing 1: dose 2 on does 1
* $H_0 : \mu = `r m_dose2`$   
* $H_a : \mu > `r m_dose2`$   
When tested under 95% confidence level, we got p-value `r pv_d2`, which is less than $\alpha = 0.05$ too. Thus we reject the null hypotheis again.  

## Conclusions
From above analysis, concusions can be made under the following assumptions:  
* Samples in ToothGrowth dataset are randomly drawn from the same population, which means the 30 samples are independent and identically distributed (iid)   
* Although with limited amount of samples in each subset, we assume the data are normally distributed  

* Conclusion 1: With 90% confidence level, the OJ has better effect than VC on the growth of tooth.  
* Conclusion 2: With 95% confidence level, dose 2 has greater effect than dose 1. The same result is observed for dose 1 on dose 0.5.  

## Appendix: source code for the report
```{r sourcecode, echo=TRUE, results="hide"}
data(ToothGrowth)
str(ToothGrowth)
table(ToothGrowth$supp); table(ToothGrowth$dose)

#with(ToothGrowth, {
#    boxplot(len ~ dose, boxwex = 0.25, at = 1:3 - 0.2,
#        subset = supp == "VC", col = "red",
#        main = "Tooth Growth under different dose of supp",
#        xlab = "dose of supp", ylab = "tooth length",
#        xlim = c(0.5, 3.5), ylim = c(0, 35), yaxs = "i")
#    boxplot(len ~ dose, add = TRUE, boxwex = 0.25, at = 1:3 + 0.2,
#        subset = supp == "OJ", col = "orange")
#    legend(2.91, 12.1, c("VC", "OJ"), fill = c("red", "orange"))
#})

g_vc <- ToothGrowth$len[ToothGrowth$supp == "VC"]
g_oj <- ToothGrowth$len[ToothGrowth$supp == "OJ"]
n_vc <- length(g_vc); n_oj <- length(g_oj)
m_vc <- round(mean(g_vc), 2); m_oj <- round(mean(g_oj), 2)
s_vc <- round(sd(g_vc) / sqrt(n_vc), 2); s_oj <- round(sd(g_oj) / sqrt(n_oj), 2)

sum_supp <- data.frame(rbind(c("VC", n_vc, m_vc, s_vc), c("OJ", n_oj, m_oj, s_oj)))
names(sum_supp) <- c("supp", "count", "mean", "sd")
#sum_supp

t.test(g_oj, g_vc, paired=FALSE, var.equal=TRUE)$conf
t.test(g_oj, g_vc, paired=FALSE, var.equal=TRUE, conf.level = 0.9)$conf

g_dose1 <- ToothGrowth$len[ToothGrowth$dose == 0.5]
g_dose2 <- ToothGrowth$len[ToothGrowth$dose == 1]
g_dose3 <- ToothGrowth$len[ToothGrowth$dose == 2]
n_dose1 <- length(g_dose1); n_dose2 <- length(g_dose2); n_dose3 <- length(g_dose3)
m_dose1 <- round(mean(g_dose1), 2)
m_dose2 <- round(mean(g_dose2), 2)
m_dose3 <- round(mean(g_dose3), 2)
s_dose1 <- round(sd(g_dose1) / sqrt(n_dose1), 2)
s_dose2 <- round(sd(g_dose2) / sqrt(n_dose2), 2)
s_dose3 <- round(sd(g_dose3) / sqrt(n_dose3), 2)

sum_dose <- data.frame(rbind(c(0.5, n_dose1, m_dose1, s_dose1), c(1, n_dose2, m_dose2, s_dose2), c(2, n_dose3, m_dose3, s_dose3)))
names(sum_dose) <- c("dose", "count", "mean", "sd")
#sum_dose

ht_d1 <- t.test(g_dose2, g_dose1, paired=FALSE, var.equal=TRUE, alternative="greater")
al_d1 <- ht_d1$alternative; pv_d1 <- ht_d1$p.value
ht_d2 <- t.test(g_dose3, g_dose2, paired=FALSE, var.equal=TRUE, alternative="greater")
al_d2 <- ht_d2$alternative; pv_d2 <- ht_d2$p.value
#ht_d1; ht_d2
```

